---
title: "Design-aware region and block detection for cross-array longitudinal EWAS"
author: "Katarzyna Kamieniecka, Samantha McLean, Krzysztof Poterlowicz"
date: "`r Sys.Date()`"
package: crossarrayEWAS
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Design-aware region and block detection for cross-array longitudinal EWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)

# The class entry points are guarded: SummarizedExperiment and GenomicRanges
# are Suggests, so the vignette still builds where they are absent.
has_bioc <- all(vapply(
  c("SummarizedExperiment", "GenomicRanges", "IRanges", "S4Vectors"),
  requireNamespace, logical(1), quietly = TRUE))
```

# What this package is for

A cohort assembled over a decade is usually assembled over two array
generations as well: earlier visits on HumanMethylation450, later ones on
MethylationEPIC. Array type is then not a nuisance variable that happens to be
present, it is a variable that moves with calendar time and therefore with
follow-up, treatment duration and often with the exposure itself. Regressing
out an array indicator does not repair that, because the indicator and the
contrast of interest are partly the same column.

`crossarrayEWAS` takes the other route: the probe-level effect is identified
**within subject**, so anything constant within a subject's visit series --
subject identity, and in a cohort where each subject sits on one array
generation, array type and chip as well -- is absorbed by the transformation
rather than modelled. What is left is estimated once, grouped by observed
co-methylation, and turned into regions and long-range blocks by two
estimators that place boundaries rather than threshold a smooth curve.

<!-- DRAFT for the author. Numbers in brackets are measured from
results/GSE237561/ (run_config.json, logs/harmonise.log, method_comparison.csv);
the judgements around them are not, and are yours to confirm or overwrite. -->

The design this was written for is a longitudinal cohort, sampled repeatedly
within subject over a recruitment window long enough that the array generation
changed mid-study: earlier subjects assayed on 450K, later subjects on EPIC,
and no subject spanning the two. [AUTHOR: confirm this is the design you want
stated, and whether to name the validation cohort here -- the package ships no
study data, and the cohort behind the manuscript (126 samples from 38
subjects, 92 arrays on 450K and 34 on EPIC) is described there rather than
here.] What makes that split non-exchangeable with exposure is that one
variable decides both: subjects recruited later sit on EPIC *and* have longer
treatment duration, so an array indicator and the contrast of interest are
partly the same column. [AUTHOR: confirm -- in the validation cohort neither a
subject nor a chip spans the two generations, so cohort is nested inside array
generation rather than crossed with it, and a batch correction taken on cohort
is not identifiable there.] Identifying the effect within subject is the
answer to that, and the cost is explicit: anything constant across a subject's
visits, including any between-subject contrast, cannot be estimated at all.
[AUTHOR: confirm you want the cost stated this plainly here rather than in the
discussion.]

```{r load}
library(crossarrayEWAS)
```

# A simulated two-array cohort

No study data ships with the package, so this vignette simulates a cohort with
the structure the estimators are designed for: subjects followed over an
unbalanced number of visits, a continuous exposure that varies **within**
subject, two array generations nested in subject, probes spaced in
co-methylated clusters, and two planted signals -- a short region and a long
low-amplitude block.

```{r simulate-design}
set.seed(20240607)

n_sub   <- 24L
visits  <- sample(2:4, n_sub, replace = TRUE)
subject <- rep(sprintf("S%02d", seq_len(n_sub)), visits)
n       <- length(subject)

# array generation is a property of the subject, not of the visit: this is the
# confounding the within-subject transform absorbs
array_type <- rep(ifelse(seq_len(n_sub) <= 14L, "450K", "EPIC"), visits)

# exposure varies between visits of the same subject, which is what makes the
# longitudinal contrast identified at all
exposure <- round(runif(n, 0, 9), 3)
covars   <- cbind(cell_fraction = rnorm(n, 0.20, 0.05),
                  smoking_score = rnorm(n, -1.0, 0.40))

c(subjects = n_sub, samples = n, on_450K = sum(array_type == "450K"))
```

Probe positions are drawn as tight clusters separated by wide gaps, which is
what `comethylation_clusters()` looks for and roughly what a methylation array
gives you.

```{r simulate-probes}
n_clust <- 640L
sizes   <- sample(2:5, n_clust, replace = TRUE)
pos     <- integer(0)
p       <- 1000000L
for (k in seq_len(n_clust)) {
  within <- c(0L, cumsum(sample(80:600, sizes[k] - 1L, replace = TRUE)))
  pos <- c(pos, p + within)
  p   <- p + max(within) + sample(3000:20000, 1L)
}
n_probes   <- length(pos)
chrom      <- rep("chr21", n_probes)
cluster_of <- rep(seq_len(n_clust), sizes)   # the truth, for comparison below
c(probes = n_probes, planted_clusters = n_clust)
```

M-values are assembled from the terms the model assumes: a probe baseline, a
subject intercept, an array offset, a per-cluster latent component that makes
neighbouring probes co-methylated, independent noise, and the exposure effect.

```{r simulate-mvalues}
sub_idx <- match(subject, unique(subject))

baseline    <- rnorm(n_probes, 0, 1.5)
subject_eff <- matrix(rnorm(n_probes * n_sub, 0, 0.30), n_probes, n_sub)
array_off   <- rnorm(n_probes, 0, 0.40)
latent      <- matrix(rnorm(n_clust * n, 0, 0.25), n_clust, n)[cluster_of, ]

# planted signal: one short region, one long shallow block
beta_true <- numeric(n_probes)
region_probes <- which(cluster_of %in% 300:301)
block_probes  <- which(cluster_of %in% 420:470)
beta_true[region_probes] <- 0.30
beta_true[block_probes]  <- 0.06

M <- baseline +
  subject_eff[, sub_idx] +
  outer(array_off, as.numeric(array_type == "EPIC")) +
  latent +
  matrix(rnorm(n_probes * n, 0, 0.20), n_probes, n) +
  outer(beta_true, exposure - mean(exposure))

rownames(M) <- sprintf("cg%07d", seq_len(n_probes))
colnames(M) <- sprintf("%s_v%d", subject,
                       unlist(lapply(visits, seq_len), use.names = FALSE))
dim(M)
```

The package takes a plain matrix of M-values with probes in rows plus parallel
vectors for the design. Nothing here needs the IDATs or an array manifest: the
harmonisation stage that builds this matrix from both array generations is a
separate step in the pipeline, and `read_f64()` is what loads the matrix it
writes.

[AUTHOR: if you would rather the vignette read a harmonised matrix from disk
than simulate one, this is the chunk to replace.]

# Probe-level effects, fitted within subject

`fit_within()` does the whole transform: it demeans the design and the
M-values by subject, drops design columns with no within-subject variation,
solves once for all probes, and moderates the residual variances towards their
common trend.

```{r fit-within}
fit <- fit_within(M, exposure, subject, covars)
str(fit)
```

The reported degrees of freedom are what the within transform leaves, not the
sample count. A subject with a single visit contributes nothing to this
contrast and is carried but not informative.

```{r fit-within-recovery}
# effects are recovered where they were planted, and not elsewhere
where <- ifelse(beta_true > 0.1, "region",
                ifelse(beta_true > 0, "block", "null"))
round(tapply(abs(fit$z), where, median), 3)
```

<!-- DRAFT for the author: the defaults are measured (bin/04_dmr_ml.R
--var-method, default "limma"), the recommendation is yours. -->

`var_method` chooses how the per-probe residual variance is moderated before
the z statistics are formed. The default `"limma"` calls `limma::squeezeVar()`,
the reference implementation of that model -- maintained, well behaved in the
degenerate cases, and already a dependency of the harmonisation stage. `"mom"`
is the method-of-moments fit ported from the Python twin of this core, and it
exists so the two implementations can be compared arithmetically on identical
input. [AUTHOR: confirm the rule you want readers to follow -- the stage
drivers default to `"limma"` and the published analysis was run with it, so
`"mom"` is for reproducing the Python path or for a deployment without the
Bioconductor stack, and should be named in the methods if a reported result
uses it.]

# Co-methylation clusters

A region should not be allowed to run across probes that are not correlated,
so probes are grouped by their **residual** co-methylation, after the exposure
effect has been projected out. This is the idiom the stage-04 driver uses, and
`pinv()` is exported so it can be reproduced here.

```{r clusters}
resid <- demean_rows_by_group(M, subject)
Xw    <- demean_by_group(cbind(exposure, covars), subject)
Xw    <- Xw[, apply(abs(Xw), 2, max) > 1e-9, drop = FALSE]
resid <- resid - (resid %*% Xw %*% pinv(crossprod(Xw))) %*% t(Xw)

cl <- comethylation_clusters(rep(21L, n_probes), pos, resid,
                             max_gap = 1000L, rho_min = 0.30)
n_cl <- if (max(cl) >= 0) max(cl) + 1L else 0L
c(clusters = n_cl, clustered_probes = sum(cl >= 0), singletons = sum(cl < 0))
```

Cluster labels are 0-based and `-1` marks a probe that joined nothing, which is
the convention the Python implementation of the same estimator uses.

# Regions

Two steps: a distance-decaying total-variation penalty, then segmentation of
the denoised track. The penalty falls off with the gap between probes, so a
wide gap is cheap to jump and a tight pair is not, and the region boundaries
are the jumps in the piecewise-constant solution rather than crossings of a
threshold.

```{r regions}
w     <- 1 / fit$se^2
lam   <- distance_penalty(pos, lam0 = 0.5, decay_bp = 1000, w = w)
theta <- tv_denoise(fit$beta, w, lam, n_iter = 3000L)

cl_arg <- ifelse(cl >= 0, cl, -seq_along(cl))
obs <- call_regions(chrom, pos, fit$beta, fit$se, theta,
                    min_probes = 3L, min_effect = 0.05, cluster = cl_arg)
reg <- obs$table
c(regions = nrow(reg), max_abs_z = round(obs$max_abs_z, 2))
head(reg[order(-abs(reg$z)),
         c("chr", "start", "end", "n_probes", "effect_M", "z")], 5)
```

`min_effect` is on the M-value scale and is a scientific choice, not a tuning
knob. [AUTHOR: the number and the reasoning are yours; what follows is a draft in
the units the run used.] In the clozapine analysis `min_effect` was 0.05 on
the M-value scale per 100 days of treatment: a segment has to move at least
that much to be called, whatever its z. At mid-range methylation that is about
0.9 percentage points of beta per 100 days -- roughly 3 points over a year and
6 over two, falling to about 2 and 4 points at beta 0.2 or 0.8 [AUTHOR:
confirm you want these conversions quoted, and at which beta]. [AUTHOR: give
the reason the floor sits there -- assay reproducibility, residual
cell-composition effect, or the smallest change you would act on -- and say
whether you hold it fixed across analyses or set it per exposure.]

`lam0` is cross-validated over held-out **subjects** in the pipeline rather
than fixed as it is here. `fold_assign()` is the split it uses: deterministic
and language-independent, so the R and Python drivers agree on it.

```{r folds}
folds <- fold_assign(subject, n_folds = 5L, seed = 0)
lengths(folds)          # subjects per fold, not samples
```

`within_df()` is the cheap test for whether a fold or a subsample is estimable
before paying for the fit: it returns the residual degrees of freedom the
within transform would leave, and zero when the exposure is annihilated
altogether.

```{r within-df}
held_out <- subject %in% folds[[1]]
c(all = within_df(exposure, subject, covars),
  train = within_df(exposure[!held_out], subject[!held_out],
                    covars[!held_out, , drop = FALSE]))
```

# Working from a SummarizedExperiment

Everything above is matrices and parallel vectors, which is what the pipeline
stages pass around. If your data is already a `SummarizedExperiment` — or a
`GenomicRatioSet`, which extends one — four functions take you in and out
without unpacking it by hand: `sort_probes()`, `probe_coords()`,
`fit_within_se()` and `as_granges()`.

```{r se-in, eval = has_bioc}
se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(M = M),
  rowRanges = GenomicRanges::GRanges(chrom, IRanges::IRanges(pos, width = 2L),
                                     cluster = cl),
  colData = data.frame(exposure = exposure, subject = subject, covars,
                       array_type = array_type))

se  <- sort_probes(se)          # genomic order, once, before the fit
co  <- probe_coords(se)
fse <- fit_within_se(se, "exposure", "subject",
                     covars = c("cell_fraction", "smoking_score"))
identical(round(fse$z[names(fit$z)], 10), round(fit$z, 10))
```

The fit is the same code: `fit_within_se()` only resolves the object, and the
identity above is the point of it. Two things it does that a hand-written
extraction usually forgets — it orders nothing silently, which is why
`sort_probes()` is a separate call, and it warns when the assay looks like
beta values, since the region and block callers assume the M scale.

Region and block calls come back as data frames because that is what the
pipeline writes to CSV. `as_granges()` turns either into a `GRanges`, keeping
`max_abs_z` in `metadata()` because that is the value a permutation null is
compared against.

```{r se-out, eval = has_bioc}
gr <- as_granges(obs)
gr[order(-abs(GenomicRanges::mcols(gr)$z))][1:3]
S4Vectors::metadata(gr)$max_abs_z
```

`SummarizedExperiment` and `GenomicRanges` are `Suggests`, not `Imports`: the
numerical core is base R plus limma so the pipeline can source it in a conda
environment with no Bioconductor stack, and the conda build of this package
inherits that. That is also why these are plain functions rather than S4
methods on `fit_within()` — `setMethod()` would need the generic's package at
install time.

# Resampling subjects, not samples

Both null constructions resample at the subject level, so a subject's visit
series is never split across a fold or a permutation.
`within_subject_permutation()` permutes exposure *inside* each subject, which
leaves the between-subject distribution intact and destroys only the
within-subject contrast -- the thing being tested.

```{r permutation}
perm <- within_subject_permutation(exposure, subject)
identical(sort(exposure[subject == "S01"]), sort(perm[subject == "S01"]))

fit_perm <- fit_within(M, perm, subject, covars)
c(observed = round(max(abs(fit$z)), 2),
  permuted = round(max(abs(fit_perm$z)), 2))
```

`stability_selection()` refits on subject-level subsamples and reports how
often each region survives. The driver default is 100 resamples; this runs 8 to
keep the vignette quick.

The callback is handed a set of **subject identifiers**, not column indices,
and returns the keys it re-discovered. Boundaries move between subsamples, so
an observed region counts as re-discovered when the intervals overlap rather
than when the endpoints match -- the same convention the stage-04 driver uses.

```{r stability}
keys <- sprintf("%s:%d-%d", reg$chr, reg$start, reg$end)

fit_subset <- function(sub) {
  sel <- subject %in% sub
  if (within_df(exposure[sel], subject[sel],
                covars[sel, , drop = FALSE]) <= 0)
    return(character(0))    # this subsample cannot estimate the contrast
  f  <- fit_within(M[, sel, drop = FALSE], exposure[sel], subject[sel],
                   covars[sel, , drop = FALSE])
  th <- tv_denoise(f$beta, 1 / f$se^2, lam, n_iter = 500L)
  r  <- call_regions(chrom, pos, f$beta, f$se, th, min_probes = 3L,
                     min_effect = 0.05, cluster = cl_arg)$table
  if (!nrow(r)) return(character(0))
  keys[vapply(seq_len(nrow(reg)), function(i)
    any(r$chr == reg$chr[i] & r$start <= reg$end[i] & reg$start[i] <= r$end),
    logical(1))]
}

set.seed(1L)
ss <- stability_selection(fit_subset, subject, n_boot = 8L, frac = 0.5)
reg$stability <- ifelse(keys %in% names(ss$freq), ss$freq[keys], 0)
reg$stability[is.na(reg$stability)] <- 0
head(reg[order(-reg$stability, -abs(reg$z)),
         c("chr", "start", "end", "n_probes", "z", "stability")], 5)
```

[AUTHOR: the threshold and its use are yours; draft follows.] The published
run resampled 50 subject subsamples at half the cohort; the eight used above
only keep the example fast [AUTHOR: confirm 50 is the number you report].
Selection frequency is carried as a column beside z rather than used as a
filter, so a region that survives the permutation null but is unstable across
subsamples stays visible and is read as exactly that. [AUTHOR: confirm, or
replace with the threshold you filter at and say which -- 0.6 and 0.8 are both
conventional in the stability-selection literature, and the choice changes
what "reported" means in the paper.]

# Long-range blocks

Blocks are called at a coarser resolution: clusters are collapsed to their mean
profile, effects are refitted at cluster level, and a three-state hidden
semi-Markov model runs along the chromosome with transition probabilities that
decay with genomic distance.

```{r collapse}
runs <- split(which(cl >= 0), cl[cl >= 0])

Mc     <- t(vapply(runs, function(rows) colMeans(M[rows, , drop = FALSE]),
                   numeric(ncol(M))))
cstart <- vapply(runs, function(rows) min(pos[rows]), numeric(1))
cend   <- vapply(runs, function(rows) max(pos[rows]), numeric(1))
cmid   <- (cstart + cend) / 2
cchr   <- rep("chr21", nrow(Mc))

o <- order(cmid)
Mc <- Mc[o, , drop = FALSE]
cstart <- cstart[o]; cend <- cend[o]; cmid <- cmid[o]
dim(Mc)
```

```{r hsmm}
fit_cl <- fit_within(Mc, exposure, subject, covars)
hs <- fit_block_hsmm(fit_cl$beta, fit_cl$se, cmid, cchr,
                     length_scale = 250000, seed = 1L)
labs <- state_labels(hs$neutral)
data.frame(state = labs, mu = round(hs$mu, 4), sigma = round(hs$sigma, 4),
           pi = round(hs$pi, 3))
```

`hs$neutral` is the column of the posterior holding the state pinned at
`mu = 0`, and it is **not** always the middle one: if every cluster effect
falls on one side of zero that column sorts to an end, and only one direction
can then be called. Pass it to `call_blocks()` rather than assuming the middle
column, or every block comes back mislabelled.

```{r blocks}
bl <- call_blocks(cchr, cstart, cend, hs$posterior, min_post = 0.80,
                  min_clusters = 3L, neutral = hs$neutral)
if (nrow(bl)) bl[order(-bl$width), ][seq_len(min(5L, nrow(bl))), ] else bl
```

```{r caption, include = FALSE}
# chunk options are evaluated before the chunk body, so the caption is set here
cap <- paste("Cluster-level effects along the chromosome. Shaded span:",
             "the planted block. Bars: the blocks called.")
```

```{r block-plot, fig.cap = cap}
planted <- range(pos[block_probes])
plot(cmid / 1e6, fit_cl$beta, pch = 16, cex = 0.5, col = "grey40",
     xlab = "position (Mb)", ylab = "effect (M-value per exposure unit)")
rect(planted[1] / 1e6, par("usr")[3], planted[2] / 1e6, par("usr")[4],
     col = "#1b9e7733", border = NA)
abline(h = 0, col = "grey70")
if (nrow(bl))
  segments(bl$start / 1e6, 0, bl$end / 1e6, 0, col = "#d95f02", lwd = 4)
```

[AUTHOR: the interpretation is yours; draft follows.] A region and a block
are not two attempts at the same quantity. Regions are called on every
harmonised probe at the resolution of a few probes -- median 84 bp over 3
probes in the published run -- and locate a focal effect. Blocks are called on
open-sea probes only, at cluster resolution, and in the same run returned 8
blocks of median 62 kb against 207 regions [AUTHOR: confirm you want the run's
geometry quoted here, or substitute your own]. A block is therefore evidence
of a shallow shift across a domain that is largely unmeasured between
clusters, and a region is evidence of a local one; the two can coexist,
and neither subsumes the other. [AUTHOR: say what you do when they disagree --
a block containing no region, or a strong region inside a neutral block. Both
defensible readings need stating explicitly: treat them as independent lines
of evidence and report both, or require a region inside a block before calling
the block interpretable. A reviewer will read the results against whichever
you choose.]

# Relationship to the pipeline and the Galaxy suite

These functions are the numerical core of a Nextflow pipeline whose earlier
stages harmonise 450K and EPIC intensities, and of a Galaxy tool suite that
wraps the same stage drivers. The pipeline is
[kkamieniecka/ewas-crossarray-harmonise](https://github.com/kkamieniecka/ewas-crossarray-harmonise),
which generates this package from its numerical core, and the wrappers are
published as a Tool Shed-installable suite at
[bioinfbrad/ewas-crossarray-harmonise](https://github.com/bioinfbrad/ewas-crossarray-harmonise);
both extend the EWASGalaxy tool suite (Murat *et al.*, *bioRxiv* 2019,
doi:10.1101/553784). [AUTHOR: confirm the links, and add the manuscript once
it has a DOI -- until then either cite it as in preparation or leave it out.
If you tag a release and mint a Zenodo DOI, cite that here as well, since the
manuscript's availability section promises one.]

# Session info

```{r session}
sessionInfo()
```
